---
title: "Module-4"
author: "Nicole Merullo"
date: "2023-09-10"
output: html_document
---

```{r matrices, echo=TRUE}
m <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2)
m
```

## Creating a Matrix
It's like having a vector within a vector. Can only have one class of data stored in it.
nrow is how many rows
ncol is how many columns
They fill up by the column first until the first column is filled then it moves over to the next.

```{r By Row False, echo=TRUE}
m <- matrix (data = c(1, 2, 3, 4, 5 ,6), nrow = 2, ncol = 3, byrow = FALSE)
m
```

## By Row False
The byrow argument is automatically set to false, so this matrix fills in its columns just like the last one did

```{r By Row True, echo=TRUE}
m <- matrix (data = c(1, 2, 3, 4, 5 ,6), nrow = 2, ncol = 3, byrow = TRUE)
m
```

## By Row True
If I change the argument to true, then it fills in the data by row instead of by column

```{r binding, echo=TRUE}
v1 <- c(1, 2, 3, 4)
v2 <- c(6, 7, 8, 9)
m1 <- rbind(v1, v2)
m1
m2 <- cbind(v1, v2)
m2
```

## Binding
With existing vectors- you can combine into matrices using rbind (by row, like having byrow = TRUE) or cbind (by column, like having byrow = FALSE)

```{r metadata, echo=TRUE}
class(m1)
dim(m1)
class(m2)
dim(m2)
colnames(m2)
rownames(m2)
```

## Metadata
Using these commands I can find out more information about the matrices. 
Note that dim() returns the dimensions of a matrix as vector and the order is row by column number.
Column names are present because I bound two vectors with defined names and cbind() was used. 