---
title: "Module-4"
author: "Nicole Merullo"
date: "2023-09-10"
output: html_document
---

```{r matrices, echo=TRUE}
m <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2)
m
```

## Creating a Matrix
It's like having a vector within a vector. Can only have one class of data stored in it.
nrow is how many rows
ncol is how many columns
They fill up by the column first until the first column is filled then it moves over to the next.

```{r By Row False, echo=TRUE}
m <- matrix (data = c(1, 2, 3, 4, 5 ,6), nrow = 2, ncol = 3, byrow = FALSE)
m
```

## By Row False
The byrow argument is automatically set to false, so this matrix fills in its columns just like the last one did

```{r By Row True, echo=TRUE}
m <- matrix (data = c(1, 2, 3, 4, 5 ,6), nrow = 2, ncol = 3, byrow = TRUE)
m
```

## By Row True
If I change the argument to true, then it fills in the data by row instead of by column

```{r binding, echo=TRUE}
v1 <- c(1, 2, 3, 4)
v2 <- c(6, 7, 8, 9)
m1 <- rbind(v1, v2)
m1
m2 <- cbind(v1, v2)
m2
```

## Binding
With existing vectors- you can combine into matrices using rbind (by row, like having byrow = TRUE) or cbind (by column, like having byrow = FALSE)

```{r metadata, echo=TRUE}
class(m1)
dim(m1)
class(m2)
dim(m2)
colnames(m2)
rownames(m2)
```

## Metadata
Using these commands I can find out more information about the matrices. 
Note that dim() returns the dimensions of a matrix as vector and the order is row by column number.
Column names are present because I bound two vectors with defined names and cbind() was used.

```{r structure, echo=TRUE}
str(m2)
```

## Structure
The str() command provides a lot of details about the object.

```{r arrays, echo=TRUE}
a <- array(data = 1:90, dim = c(5, 6, 3))
a
```

## Arrays
This created an array with data all integers 1 through 90 arranged in 3 levels, 5 rows by 6 columns
This is a 3 dimensional array (there is a third dimension beyond rows and columns)
Arrays can only have one class of data

```{r Subsetting, echo=TRUE}
v <- 1:100
v
v[1:15] #selects first 15 elements
v[c(2, 4, 6, 8 ,10)] #creates vector of selected elements
```

## Subsetting
These are some basic ways of getting a few elements from a large vector

```{r Subsetting2, echo=TRUE}
v <- 101:200
v[seq(from = 1, to = 100, by = 2)]
```

#Subsetting with a Function or Calculation
This subset selected integers counting by 2s, set by the dimensions in the from and to arguments