---
title: "Module-4"
author: "Nicole Merullo"
date: "2023-09-10"
output: 
  html_document:
    theme: yeti
    highlight: NULL
---
## Creating a Matrix

```{r matrices, echo=TRUE}
m <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2)
m
```

It's like having a vector within a vector. Can only have one class of data stored in it.
nrow is how many rows
ncol is how many columns
They fill up by the column first until the first column is filled then it moves over to the next.

## By Row False

```{r By Row False, echo=TRUE}
m <- matrix (data = c(1, 2, 3, 4, 5 ,6), nrow = 2, ncol = 3, byrow = FALSE)
m
```

The byrow argument is automatically set to false, so this matrix fills in its columns just like the last one did

## By Row True

```{r By Row True, echo=TRUE}
m <- matrix (data = c(1, 2, 3, 4, 5 ,6), nrow = 2, ncol = 3, byrow = TRUE)
m
```

If I change the argument to true, then it fills in the data by row instead of by column

## Binding

```{r binding, echo=TRUE}
v1 <- c(1, 2, 3, 4)
v2 <- c(6, 7, 8, 9)
m1 <- rbind(v1, v2)
m1
m2 <- cbind(v1, v2)
m2
```

With existing vectors- you can combine into matrices using rbind (by row, like having byrow = TRUE) or cbind (by column, like having byrow = FALSE)

## Metadata

```{r metadata, echo=TRUE}
class(m1)
dim(m1)
class(m2)
dim(m2)
colnames(m2)
rownames(m2)
```

Using these commands I can find out more information about the matrices. 
Note that dim() returns the dimensions of a matrix as vector and the order is row by column number.
Column names are present because I bound two vectors with defined names and cbind() was used.

## Structure

```{r structure, echo=TRUE}
str(m2)
```

The str() command provides a lot of details about the object.

## Arrays

```{r arrays, echo=TRUE}
a <- array(data = 1:90, dim = c(5, 6, 3))
a
```

This created an array with data all integers 1 through 90 arranged in 3 levels, 5 rows by 6 columns
This is a 3 dimensional array (there is a third dimension beyond rows and columns)
Arrays can only have one class of data

## Subsetting

```{r Subsetting, echo=TRUE}
v <- 1:100
v
v[1:15] #selects first 15 elements
v[c(2, 4, 6, 8 ,10)] #creates vector of selected elements
```

These are some basic ways of getting a few elements from a large vector

## Subsetting with a Function or Calculation

```{r Subsetting2, echo=TRUE}
v <- 101:200
v[seq(from = 1, to = 100, by = 2)]
```

This subset selected integers counting by 2s, set by the dimensions in the from and to arguments

## Challenge 1

```{r Challenge1, echo=TRUE}
u <- c("stately", "plump", "buck", "mulligan", "came", "from", "the", "stairhead", "bearing", "a", "bowl", "of", "lather", "on", "which", "a", "mirror", "and", "a", "razor", "lay", "crossed")
u
u[seq(from = 1, to = 22, by = 3)]
```

Trying again with the str_split() command
leaving the pattern argument blank (just closed quotation marks) would break it up by character, adding the space broke it up by word. 

```{r Challenge1 take 2, echo=TRUE}
u <- c("Stately plump Buck Mulligan came from the stairhead bearing a bowl of lather on which a mirror and a razor lay crossed")
u
require(stringr)
u <- str_split_1(u, " ")
u[seq(from = 1, to = 22, by = 3)]
```

## Challenge 2

```{r Challenge2, echo=TRUE}
m <- matrix(data = 1:80, nrow = 8, ncol = 10, byrow = FALSE)
m
```

### extracting 2nd, 3rd, and 6th columns 

```{r Challenge2 part 1, echo=TRUE}
x <- c(m[, 2], m[, 3], m[, 6])
x
```

### extracting 6th to 8th rows

```{r Challenge2 part 2, echo=TRUE}
x <- c(m[6, ], m[7, ], m[8, ])
x
```

### extracting the elements from row 2, column 2 to row 6, column 9

```{r Challenge2 part 3, echo=TRUE}
x <- c(m[2, ], m[, 2:6], m[, 9])
x
```

## Challenge 3
Create a 4-dimensional 400 element array (5x5x4x4) named a consisting of numbers 400 to 1 (descending)
Subsetting this array requires 4 coordinates. Each one corresponds to a location/dimension. It's easiest to work backwards so the farthest right dimension working left to figure out how it gets a result. When a dimension is missing, it results any of the 4 options. 

```{r Challenge3, echo=TRUE}
a <- array(data = 400:1, dim = c(5, 5, 4, 4))
a
a[1, 1, 1, 2]
a[2, 3, 2, ]
a[1:5, 1:5, 3, 3]
```

## Overwriting
Use m as defined above
Note that as soon as some of them are characters they ALL have to become characters because matrices can only have one class of data.

```{r Overwriting, echo=TRUE}
m
m[7, 1] <- 564
m[, 8] <- 2
m[2:5, 4:8] <- 1
m[2:5, 4:8] <- c(20, 19, 18, 17)
m[2:5, 4:8] <- matrix(data = c(20:1), nrow = 4, ncol = 5, byrow = TRUE)
m[, 8] <- c("a", "b")
m
```

## Lists and Data Frames
Below are different commands and ways to extract details from lists

```{r Lists, echo=TRUE}
s <- c("this", "is", "a", "vector", "of", "strings")  #vector
m <- matrix(data = 1:40, nrow = 5, ncol = 8)  #matrix
b <- FALSE  #logical statement
l <- list(s, m, b)  # combines into list
l
l[[2]] #just looking at a part of a list, in this case the matrix
l[[2]][2, 6] #extending the notation to normal subsetting notation
l[1:2] #list slicing
l[c(1, 3)] #list slicing again but this time just 1 and 3
str(l)
names(l) <- c("string", "matrix", "logical")
names(l)
l$string
l$matrix[3, 5]
```

## Challenge 5
Create lists within lists to reconstruct a primate phylogeny

```{r Challenge 5, echo=TRUE}
Lemuroidea <- c("Cheirogaleidae", "Lepilemuridae", "Indriidae", "Lemuridae",
    "Daubentoniidae")
Lorisoidea <- c("Lorisidae", "Galagidae")
Strepsirhini <- list(Lemuroidea, Lorisoidea)
names(Strepsirhini) <- list("Lemuroidea", "Lorisoidea")
Strepsirhini
Tarsioidea <- c("Tarsiidae")
Catarrhini <- c("Cercopithecidae", "Hylobatidae", "Hominidae")
Platyrrhini <- c("Cebidae", "Atelidae", "Pitheciidae")
Anthropoidea <- list(Platyrrhini, Catarrhini)
names(Anthropoidea) <- list("Platyrrhini", "Catarrhini")
Haplorhini <- list(Anthropoidea, Tarsioidea)
names(Haplorhini) <- list("Anthropoidea", "Tarsioidea")
Primates <- list(Haplorhini, Strepsirhini)
names(Primates) <- list("Haplorhini", "Stepsirhini")
Primates
```

## Data Frames
Rows are cases/observations and Columns are variables

```{r Data Frame Example, echo=TRUE}
df <- data.frame(firstName = c("Nitin", "Silvy", "Ingrid", "Claire"), major = c("Biology (CM)",
    "Anthropology (Bio)", "Anthropology (Medical)", "Biology (EEB)"), sex = c("M",
    "F", "F", "F"), yearInProgram = c(1, 2, 2, 3))
df
```

## Reading a CSV
Select file name and location on computer. Line separator is comma. There are headers (true). We are only printing select columns so the row is left blank and then a vector is entered for the non continuous columns.
Structure tells us the different columns and the first few observations of each one. Also the class.

```{r Reading a CSV, echo-TRUE}
df <- read.csv("~/Desktop/random-people.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)
df[, c(1, 3, 4, 11, 12)]
```

## Data Frame Commands
Structure tells us the different columns and the first few observations of each one. Also the class.
```{r Daa Frame Commands, echo-TRUE}
str(df)
df[, 3] #all first names (data from column 3)
str(df[,3]) #returns a vector of the data from column 3
df$name.last #another way of finding the observations of a column name/variable (if you know the name)
str(df$name.last) #returns a vector of the last names
df[3] #extracts data from column 3 as data frame
str(df[3]) #proving df[3] is a data frame structure
df["name.last"]
str(df["name.last"]) #same thing as before - proves that it comes out as a data frame
```

##Modifying Data Frames
use rbind and cbind to add rows and columns to data frames

```{r Modfiying Data Frames, echo-TRUE}
df <- cbind(df, id = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16,
    17, 18, 19, 20))
df <- cbind(df, school = c("BU", "BU", "Harvard", "Harvard", "BU", "Boston College",
    "MIT", "BU", "BU", "UMASS Boston", "Harvard", "BU", "Boston College", "BU",
    "Harvard", "MIT", "Harvard", "BU", "UMASS Boston", "Harvard"))
df
```

Or you can extend the data frame using $ but this will etend it as a character vector as opposed to a factor

```{r Extending a data frame, echo-TRUE}
df$school <- c("BU", "BU", "Harvard", "Harvard", "BU", "Boston College", "MIT",
    "BU", "BU", "UMASS Boston", "Harvard", "BU", "Boston College", "BU", "Harvard",
    "MIT", "Harvard", "BU", "UMASS Boston", "Harvard")
```

Factors are an atomic data class. They are unique because R assigns eahch unique character strings to an integer so something like "male" becomes 2 and "female" becomes 1. 

## Filtering
Using logical vectors to subset data frames

```{r Filtering1, echo-TRUE}
new_df <- df[df$school =="BU", ]
new_df
```

And if we only want to see a few columns when we do this, that's possible too.

```{r Filtering2, echo-TRUE}
new_df <- df[df$school =="BU", c("name.last", "name.first", "school")]
new_df
```

## Challenge 5
Store the following numbers as a 5 x 3 matrix, called “m3”: 3, 0, 1 ,23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2. Be sure to fill the matrix ROW-WISE.

```{r Challenge 5, echo-TRUE}
m3 <- matrix(data = c(3, 0, 1 ,23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2), nrow = 5, ncol = 3, byrow = TRUE)
m3
```

Coerce from Matrix to Data Frame

```{r Challenge 5-2, echo-TRUE}
m3 <- as.data.frame(m3)
m3
```

Coerce the second column to be logic-valued

```{r Challenge 5-3, echo-TRUE}
m3 [, 2] <- as.logical(m3[, 2])
m3
```

Coerce the third column to be factor valued

```{r Challenge 5-4, echo-TRUE}
m3 [, 3] <- as.factor(m3[, 3])
m3
```

Use structure to see the data type for each variable (column)

```{r Challenge 5-5, echo-TRUE}
str(m3)
```