---
title: "Module-10"
author: "Nicole Merullo"
date: "2023-10-03"
output: html_document
---
#Classical Hypothesis Testing
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(curl)
```

*text copied from module instructions for reference with personal notes added in*

## Null and Alternaitve Hypotheses

Classical or frequentist hypothesis testing (a.k.a. parametric statistics) involves formally stating a claim - the null hypothesis - which is then followed up by statistical evaluation of the null versus an alternative hypotheses. The null hypothesis is interpreted as a baseline hypothesis and is the claim that is assumed to be true. The alternative hypothesis is the conjecture that we are testing.

We need some kind of statistical evidence to reject the null hypothesis in favor of an alternative hypothesis. This evidence is, in classical frequentist approaches, some measure of how unexpected it would be for the sample to have been drawn from a given null distribution.

H0 = null hypothesis = a sample statistic shows no deviation from what is expected or neutral

HA = alternative hypothesis = a sample statistic deviates more than expected by chance from what is expected or neutral

We can test several different comparisons between H0 and HA.

HA > H0, which constitutes an "upper one-tailed test" (i.e., our sample statistic is greater than that expected under the null)

HA < H0, which constitutes an "lower one-tailed test" (i.e., our sample statistic is less than that expected under the null)

HA ≠ H0, which constitutes an "two-tailed test" (i.e., our sample statistic is different, maybe greater maybe less, than that expected under the null)

There are then four possible outcomes to our statistical decision:

Outcomes
true: H0 Decide: H0 Result: Correctly 'accept' the null
true: H0 Decide: HA Result: Falsely reject the null (Type I error)
true: HA Decide: H0 Result: Falsely accept the null (Type II error)
true: HA Decide: HA Result: Correctly reject the null

In classical frequentist (a.k.a. parametric) inference, we perform hypothesis testing by trying to minimize our probability of Type I error… we aim for having a high bar for falsely rejecting the null (e.g., for incorrectly finding an innocent person guilty). When we set a high bar for falsely rejecting the null, we lower the bar for falsely ‘accepting’ (failing to reject) the null (e.g., for concluding that a guilty person is innocent).

To do any statistical test, we typically calculate a test statistic based on our data, which we compare to some appropriate standardized sampling distribution to yield a p value.

**The p value = the probability of our obtaining a test statistic that is as high or higher than our calculated one by chance, assuming the null hypothesis is true.**

The test statistic basically summarizes the “location” of our data relative to an expected distribution based on our null model. The particular value of our test statistic is determined by both the difference between the original sample statistic and the expected null value (e.g., the difference between the mean of our sample and the expected population mean) and the standard error of the sample statistic. The value of the test statistic (i.e., the distance of the test statistic from zero) and the shape of the null distribution are the sole drivers of the smallness of the p value. The p value effectively represents the area under the sampling distribution associated with test statistic values as or more extreme than the one we observed.

We compare the p value associated with our test statistic to some significance level, α, typically 0.05 or 0.01, to determine whether we reject or fail to reject the null. If p < α, we decide that there is sufficient statistical evidence for rejecting H0.

How to calculate the p value:

1. specify the test statistic like the mean
2. specify the null distribution i.e., how the null statistic is distributed (normal, poisson etc)
3. Calculate the tail probability, i.e., the probability of obtaining a statistic (mean etc) as or more extreme than was observed assuming that null distribution. 

3 ? Basically what is the probability of getting a more extreme mean (given resampling?)

## Testing Sample Means: One Sample Z and T Tests

Let’s do an example where we try to evaluate whether the mean of a single set of observations is significantly different than expected under a null hypothesis… i.e., this is a ONE-SAMPLE test.

Suppose we have a vector describing the adult weights of vervet monkeys trapped in South Africa during the 2016 trapping season. We have the sense they are heavier than vervets we trapped in previous years, which averaged 4.9 kilograms. The mean is 5.324 kilograms. Is the mean significantly greater than our expectation?
```{r samplemeans-1}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/vervet-weights.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

Mean of vervet weights from this year:

```{r samplemeans-2}
mean(d$weight)
```

**What is H0?** Vervet mean weights are the same (Vervet weights are not greater than 4.9)
**What is HA?** Vervet mean weights are different (Vervet weights are great than 4.9)
Could be an upper tailed test or two tailed test (the latter more the way I described it, but if I had said vervet mean weights are greater this year as the HA then it would more be an upper tailed test)

Mean, SD, SEM from Sample:

```{r samplemeans-3}
mu <- 4.9 #sample mean from previous years
x <- d$weight
m <- mean(x)
s <- sd(x) #how much does the data (weights of vervets) vary from the mean
n <- length(x) #(number of vervets)
sem <- s/sqrt(n) #how reliable is this sample data in representing the population
```

Our test statistic takes a familiar form… it is effectively the standard normalized position of our sample mean in a distribution centered around the expected population mean.

And the formula:
Z = (mean of sample obs - expected mean) / (sample SD/sqrt # of sample obs)

```{r samplemeans-4}
z <- (m - mu)/sem
z
```

In this case, z is a quantile… the estimated number of standard errors of the mean away from the population mean that our sample falls.

If we then want to see if z is significant, we need to calculate the probability of seeing a deviation from the mean as high or higher than this by chance. To do this, we can use the pnorm() function. Because we have converted our sample mean to the standard normal scale, the mean= and sd= arguments of pnorm() are the defaults of 0 and 1, respectively.

Remember that pnorm returns the cumulative probability less than or equal to the given quantile

We want the probability of seeing a z this large or larger by chance.

```{r samplemeans-5}
p <- 1 - pnorm(z)
p
```

However, as noted above, our sample size from a population is typically limited. So, instead of using the normal distribution to determine the p value of our statistic, we should use the t distribution, which, as we’ve seen, has slightly fatter tails. The statistic and process is exactly the same, though, as for the normal distribution.

```{r samplemeans-6}
p <- 1 - pt(z, df = n - 1)
p
p <- pt(z, df = n - 1, lower.tail = FALSE) #lower.tail false is probability of greater than z (3 SEM away)
p #results are exactly the same
```

Shows slightly less significance than with the normal distribution but that makes sense because of the fatter tails/less certainty

R has built into it a single function, t.test(), that lets us do all this in one line. We give it our data and the expected population mean, μ, along with the kind of test we want to do.

```{r samplemeans-7}
t <- t.test(x = x, mu = mu, alternative = "greater")
t
```

CI is also returned though it shows inf as one end??

By hand:

```{r samplemeans-8}
lower <- m - qt(1 - 0.05/2, df = n - 1) * sem
upper <- m + qt(1 - 0.05/2, df = n - 1) * sem
ci <- c(lower, upper)
ci
```

By T test:
Have to use two sided arugment
```{r samplemeans-9}
t <- t.test(x = x, mu = mu, alternative = "two.sided")
ci <- t$conf.int
ci
```

Reject H0 - the vervet weights are signifcantly heavier than the previous trapping season (p<0.01)
Avg from previous season falls outside of 95% CI for this season

## Challenge 1

Adult lowland woolly monkeys are reported to have an average body weight of 7.2 kilograms. You are working with an isolated population of woolly monkeys from the Peruvian Andes that you think may be a different species from the lowland form, and you collect a sample of 15 weights from adult individuals at that site. From your sample, you calculate a mean of 6.43 kilograms and a standard deviation of 0.98 kilograms.

Perform a hypothesis test of whether body weights in your population are different from the reported average for lowland woolly monkeys by setting up a “two-tailed” hypothesis, carrying out the analysis, and interpreting the p value (assume the significance level is α=0.05). Your sample is < 30, so you should use the t distribution and do a t test. Do your calculations both by hand and using the t.test() function and confirm that they match.

```{r Challenge1-1}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall21/woolly-weights.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

```{r Challenge1-2}
mu <- 7.2 #sample mean of lowland woolly monkeys
x <- d$weight
m <- mean(x)
m
s <- sd(x) #how much does the data (weights of woolly monkeys vary from the mean
n <- length(x) #(number of woolly monkeys)
sem <- s/sqrt(n) #how reliable is this sample data in representing the population
sem
t <- (m - mu)/sem
t
p <- 1 - pt(t, df = n - 1)
p
p <- pt(t, df = n -1, lower.tail=FALSE)
p
p1 <- pt(t, df = n - 1)
p1
```

H0 is that they are the same, HA is that they are different
H0 is accepted because the p value is >0.05. The cumulative probability of getting anything other than </= -3.34 is very high and (right side) same if you do >-3.34 which is the same thing (right side). But the cumulative probability of just getting </= -3.34 (the left side) is significant.

Here is what is included in the module:

```{r Challenge1-3}
alpha <- 0.05
crit <- qt(1 - alpha/2, df = n - 1)  # identify critical values
test <- t < -crit || t > crit  # boolean test as to whether t is larger than the critical value at either tail
test <- abs(t) > crit
test
```

Basically is the t value larger than value at tails where significance is reached. It is true so it is >0.05 thus not significant??

```{r Challenge1-4}
mu <- 7.2 #sample mean of lowland woolly monkeys
x <- d$weight
m <- mean(x)
m
s <- sd(x) #how much does the data (weights of woolly monkeys vary from the mean
n <- length(x) #(number of woolly monkeys)
sem <- s/sqrt(n) #how reliable is this sample data in representing the population
sem
t.test(x=x, mu=mu, alternative = "two.sided")
```

I got very different p values in my by hand version... this one is showing significance.

Conflicting results, cannot decide on hypothesis. Cannot detemrine if these are a separate species or not.

## Comparing Sample Means: Two Sample Z and T Tests

Sometimes we want to compare two groups of measurements to one another, which boils down to a hypothesis test for the difference between two means, μ1 and μ2. **The null hypothesis is that the difference between the means is zero.**

Before getting to the appropriate test, there are a couple of things that we need to consider:

[1] How, if at all, are the two samples related to one another? Sometimes we may have PAIRED samples (e.g., the same individuals before and after some treatment) and sometimes the samples are UNPAIRED or INDEPENDENT (e.g., weights for different samples of black-and-white colobus monkeys collected in the rainy versus dry seasons).

[2] Are the variances in the two samples roughly equal or not? E.g., if we are comparing male and female heights, are the variances comparable?

### Samples with unequal variances

For the most generic case, where the two samples are independent and we cannot assume the variances of the two samples are equal, we can do what is called Welch’s t test where our test statistic is:

T = (mean of x2 = mean of x1 - mu)/(sqrt (sd1 squared/# of obs 1 + sd2 squared/# of obs 2))


## Challenge 2

Let’s do an example. Load in a file of black-and-white colobus weights and examine the str() of the file. Then, create 2 vectors, x and y, for male and female weights. Plot these in boxplots side by side and then calculate the mean, sd, and sample size for both males and females.

```{r Challenge2-1}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/colobus-weights.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
str(d)
```

vectors for male and female weight
```{r Challenge2-2}
x <- d$weight[d$sex == "male"]
y <- d$weight[d$sex == "female"] #subset then subset again
x
y
```

boxplot

```{r Challenge2-3}
par(mfrow = c(1,2))
boxplot(x, ylim = c(4.5, 8), main = "Weight (kg)", xlab = "Males")
boxplot(y, ylim = c(4.5, 8), main = "Weight (kg)", xlab = "Females")
```

determining means, sds, and ns

```{r Challenge2-4}
m1 <- mean(x)
m1
m2 <- mean(y)
m2
mu <- 0  # you could leave this out... the default argument value is 0
s1 <- sd(x)
s1
s2 <- sd(y)
s2
n1 <- length(x)
n2 <- length(y)
```

Now calculate the t stat and test the two tailed hypothesis

```{r Challenge2-5}
t <- (m2-m1-mu) / sqrt(s2^2/n2 + s1^2/n1)
t
```

and the critical values part 

```{r Challenge2-6}
alpha <- 0.05
crit <- qt(1 - alpha/2, df = n - 1)  # identify critical values
crit #basically the values at probability of 0.05
```

And now seeing if the t stat is larger than the critical values

```{r Challenge2-7}
test <- t < -crit || t > crit  # boolean test
test <- abs(t) > crit
test
```

Note that for this test, the number of degrees of freedom is calculated as:

```{r Challenge2-8}
df <- (s2^2/n2 + s1^2/n1)^2/((s2^2/n2)^2/(n2 - 1) + (s1^2/n1)^2/(n1 - 1))
df
```

Ok same thing with t tes function

```{r Challenge2-9}
t <- t.test(x=x, y=y, mu=mu, altnernative= "two.sided")
t
```

P value is very small thus is significant. We reject the null hypothesis.

## Samples with Equal Variances

There’s a simpler t statistic we can use if the variances of the two samples are more or less equal.
T = (mean of x2 - mean of x1 - mu)/ sqrt (sd of pooled sample squared(1/#of obs1 + 1/#of obs 2))
See module instructions for how to get sd of pooled sample (they have the same variance from the mean which is why we can do this)

df = n1 + n2 -2

```{r EqualVar-1}
s <- sqrt((((n1 - 1) * s1^2) + ((n2 - 1) * s2^2))/(n1 + n2 - 2))
t <- (m2 - m1 - mu)/(sqrt(s^2 * (1/n1 + 1/n2)))
t
df <- n1 + n2 - 2
df
t <- t.test(x = x, y = y, mu = 0, var.equal = TRUE, alternative = "two.sided")
t
```

Even more significance

A crude test for equality of variances is to divide the larger by the smaller and if the result is < 2, you can go ahead and used the pooled variance version of the test (which has many fewer degrees of freedom).

```{r EqualVar-2}
var(x)/var(y)
```

This is greater than 2 so they have different variances for the purposes of these tests

We can use the var.test() function to conduct an actual statistical test on the ratio of variances, which compares the ratio test statistic we just calculated to an F distribution. The F distribution is often used to model ratios of random variables and thus is useful in regression applications and, as here, for testing whether variances from two samples are different. It is dependent upon the specification of a pair of degrees of freedom values supplied as the arguments df1= and df2= (or inferred from the number of observations in each sample).

Below, the results of var.test() are saved to a variable. Calling the variable provides a brief descriptive summary.

```{r EqualVar-3}
vt <- var.test(x, y)
vt
```

## Paired Samples

For a paired sample test, the null hypothesis is that the mean of individual paired differences between the two samples (e.g., before and after) is zero.

Our test statistic is:

T = (mean of difference b/w paired samples - mu) / (sqrt (sd of differences squared / # of sample pairs))

Again, note that μ here is the expected difference between the means under the null hypothesis, which is zero, and we are dividing by the standard error of the mean for the set of differences between pairs.

## Challenge 3

Let’s play with a sample… IQs of individuals taking a certain statistics course before and after a lecture on significance testing. Load in the iqs.csv data file, look at it, plot a barchart of values before and after and construct a paired t test to evaluate the means before and after.

```{r Challenge3-1}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/iqs.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

```{r Challenge3-2}
x <- d$IQ.before - d$IQ.after
m <- mean(x)
mu <- 0  # can leave this out
s <- sd(x)
n <- length(x)
sem <- s/sqrt(n)
par(mfrow = c(1, 2))
boxplot(d$IQ.before, ylim = c(115, 145), main = "IQ", xlab = "Before")
boxplot(d$IQ.after, ylim = c(115, 145), main = "IQ", xlab = "After")
```

two tailed test

```{r Challenge3-3}
t <- (m - mu)/sem
t
alpha <- 0.05
crit <- qt(1 - alpha/2, df = n - 1)  # identify critical values
crit
test <- t < -crit || t > crit  # boolean test
test
t.test(x, df = n - 1, alternative = "two.sided")
```

NOT SIGNIFICANT

## Testing Sample Proportions: One Sample Z Test

As we have seen, the sampling distribution of sample means for independent and identically distributed random variables is roughly normal (and, as shown by the CLT, this distribution increasingly approaches normal as sample size increases). Similarly, the sampling distribution for another kind of sample statistic, the number of “successes” x out of a series of k trials is also roughly normally distributed. If the true population proportion of “successes” is π, then the sampling distribution for the proportion of successes in a sample of size n is expected to be roughly normally distributed with mean = π and standard error = sqrt(π(1−π)/n).

Let’s set up a simulation to show this…

First we create a population of 500 “1”s and 500 “0”s, i.e., where π = 0.5…

```{r ZTest-1}
pop <- c(rep(0, 500), rep(1, 500))
```

Now, we will take 1000 random samples of size n=10 from that population and calculate the proportion of “1”s in each sample…

```{r ZTest-2}
pi <- 0.5
x <- NULL
n <- 10
for (i in 1:1000) {
    x[i] <- mean(sample(pop, size = n, replace = FALSE))  # taking the mean of a bunch of 0s and 1s yields the proportion of 1s!
}
m <- mean(x)
m
s <- sd(x)
s
pop_se <- sqrt(pi * (1 - pi)/n)
pop_se  # the se is an estimate of the sd of the sampling distribution
```

The same is true if we create a population of 800 “1”s and 200 “0”s, i.e., where π = 0.8…

```{r ZTest-3}
pop <- c(rep(0, 800), rep(1, 200))
pi <- 0.8
x <- NULL
n <- 10
for (i in 1:1000) {
    x[i] <- mean(sample(pop, size = n, replace = FALSE))  # taking the mean of a bunch of 0s and 1s yields the proportion of 1s!
}
m <- mean(x)
m
s <- sd(x)
s
pop_se <- sqrt(pi * (1 - pi)/n)
pop_se  # the se is an estimate of the sd of the sampling distribution
```

This normal approximation is true as long as n is fairly large and π is not close to 0 or 1. One rule of thumb is to check that both n×π and n×(1−π) are greater than 5.

With all this in mind, we can construct Z statistics for proportions like we constructed Z and T statistics for means and test those proportions for differences from an expected value or for differences between two sample proportions. The Z statistic for proportions takes the same general form as that for means…

z = (observed statistic - expected statistic) / standard error

## Challenge 4

A neotropical ornithologist working in the western Amazon deploys 30 mist nets in a 100 hectare (ha) grid. She monitors the nets on one morning and records whether or not she captures any birds in the net (i.e., a “success” or “failure” for every net during a netting session). The following vector summarizes her netting results:

```{r Challenge4-1}
v <- c(0, 1, 1, 1, 0, 1, 0, 0, 1, 1, 0, 0, 1, 1, 0, 1, 1, 0, 0, 1, 1, 1, 0,
    1, 1, 0, 1, 0, 1, 1)
```

Her netting success over the previous three seasons suggests that she should catch birds in 80% of her nets. This season, she feels, her success rate is lower than in previous years. Does her trapping data support this hypothesis?

**What is H0?**
Netting success is the same as last year
**What is HA?**
Netting success is lower than previous years
**Are both n×π and n×(1−π) > 5?**
```{r Challenge4-2}
n <- length(v)
pi <- 0.8 #success rate of birds in nests
n*pi
n*(1-pi)
```
**Calculate z and the p value associated with z**
```{r Challenge4-3}
phat <- mean(v)
phat
z <- (phat - pi)/sqrt(pi * (1 - pi)/30)
z
```
**Calculate the 95% CI around** 
```{r Challenge4-4}
p <- pnorm(z, lower.tail = TRUE) #lower tail true because it is a lower tailed one tailed hypothesis (the rate is lower than previous years)
p
lower <- phat - qnorm(0.975) * sqrt(phat * (1 - phat)/30)
upper <- phat + qnorm(0.975) * sqrt(phat * (1 - phat)/30)
ci <- c(lower, upper)
ci
```

This approach using quantiles of the standard normal distribution is but one method of calculating CIs for proportion data, and is the CI referred to as a Wald confidence interval. Note that the CI does not include the value of π… rather, π > is greater than the upper bound of the CI, suggesting that the observed success rate is indeed lower than in previous years.

We can do the same test with a one-liner in R…

```{r Challenge4-5}
pt <- prop.test(x = sum(v), n = length(v), p = 0.8, conf.level = 0.95, correct = FALSE,
    alternative = "less")
pt
```

Note that the CI is different than we calculated based on the normal distribution… prop.test() implements a slightly different procedure for estimating the CI rather than basing this on the normal distribution and the CLT.

How should she interpret these results? Was she right that her trapping results are off this year?

*Average of this year being 60% is significantly lower than previous year*

## Comparing Sample Proportions: Two Sample Z Tests

The Z statistic for the two sample test comparing proportions is also very similar to that for comparing means.

Z = (phat2 - phat1 - pi) / sqrt (pstar(1-pstar)(1/n1+1/n2))

phat is the proportions of successes in each sample group
pi is the expected difference in proportions between sample groups (usually set to 0)
pstar is the pooled proportion
Pstar = x1 + x2 / n1 + n2

## Challenge 5

A biologist studying two species of tropical bats captures females of both species in a mist net over the course of week of nightly netting. For each species, the researcher records whether females are lactating or not. The two vectors below summarize the data for each species.

```{r Challenge5-1}
v1 <- c(1, 0, 1, 1, 0, 0, 1, 1, 0, 0, 1, 1, 1, 1, 0, 0, 1, 1, 0, 0, 1, 1, 0,
    1, 0)
v2 <- c(1, 1, 0, 0, 1, 1, 1, 1, 0, 1, 0, 1, 1, 1, 0, 1, 0, 1, 1, 1, 1, 1, 0,
    0, 1, 1, 0, 1, 1, 1)
```

Based on your mist netting data, do the species differ significantly in the proportion of lactating females during this trapping season? What are H0 and HA?

H0 is no difference (pi is zero), HA is difference (two tailed hypothesis)

```{r Challenge5-2}
phat1 <- mean(v1)
phat1
phat2 <- mean(v2)
phat2
pi <- 0
```

pstar:
Pstar = x1 + x2 / n1 + n2
```{r Challenge5-3}
pstar <- (sum(v1) + sum(v2)) / (length(v1)+length(v2))
pstar
```

z value

```{r Challenge5-4}
z <- (phat2 - phat1)/sqrt((pstar * (1 - pstar)) * (1/length(v1) + 1/length(v2)))
z
```

pnorm for p value plus crit values thing for two tailed hypothesis

```{r Challenge5-5}
p <- 1 - pnorm(z, lower.tail = TRUE) + pnorm(z, lower.tail = FALSE)
p
crit <- qnorm(1 - alpha/2)  # identify critical values
crit
test <- p < -crit || p > crit  # boolean test
test
```
False is not significant

We can use the prop.test() function to do this in one line.

```{r Challenge5-6}
pt <- prop.test(x = c(sum(v2), sum(v1)), n = c(length(v2), length(v1)), alternative = "two.sided",
    correct = FALSE)
pt
```
What are her results? Are the proportions of lactating females different across species during this trapping season?

Not significant

## Summary of Z and T Tests

Z and T tests are used to evaluate whether a given sample statistic (e.g., a mean or proportion) deviates significantly from what is expected under a null model or whether two samples statistics deviate significantly from one another.

They are used for dealing with normally distributed, continuous variables or those that can be approximated closely by the normal distribution.

-We REJECT a H0 if the p value obtained for a given Z or T test statistic is < α
-CIs for our sample statistic are calculated as mean ± T(1−α/2) or Z(1−α/2) x SEM, and we can REJECT a H0 if the (1-α) CI around does not include the expected value of the statistic
=When the sample size > 30, or when we are dealing with proportions, we use Z quantiles for calculating CIs and p values, but for sample size < 30, we use T quantiles

See module for the various formulae for the tests