---
title: "Module-4"
author: "Nicole Merullo"
date: "2023-09-10"
output: html_document
---

```{r matrices and arrays, echo=TRUE}
m <- matrix(data = c(1, 2, 3, 4), nrow = 2, ncol = 2)
m
```

## Creating a Matrix
It's like having a vector within a vector. Can only have one class of data stored in it.
nrow is how many rows
ncol is how many columns
They fill up by the column first until the first column is filled then it moves over to the next.

```{r By Row False, echo=TRUE}
m <- matrix (data = c(1, 2, 3, 4, 5 ,6), nrow = 2, ncol = 3, byrow = FALSE)
m
```

## By Row False
The byrow argument is automatically set to false, so this matrix fills in its columns just like the last one did

```{r By Row True, echo=TRUE}
m <- matrix (data = c(1, 2, 3, 4, 5 ,6), nrow = 2, ncol = 3, byrow = TRUE)
m
```

## By Row True
If I change the argument to true, then it fills in the data by row instead of by column

```{r binding, echo=TRUE}
v1 <- c(1, 2, 3, 4)
v2 <- c(6, 7, 8, 9)
m1 <- rbind(v1, v2)
m1
m2 <- cbind(v1, v2)
m2
```

