---
title: "Module-6"
author: "Nicole Merullo"
date: "2023-09-16"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
```

## Some useful procedures for exploratory data analysis for univariate data

### Challenge 1

Using summary() to get an overview of data from each column in a given data frame
Includes (for numeric data):
1. minimum
2. 25th percentile
3. median
4. mean
5. 75th percentile
6 maximum
7. count of NAs (missing values)
For Factors:
1. Count of each factor level (how many 1s, 2s, 3s)

First load in the file

```{r Challenge1-1, echo=TRUE}
f <- "~/Documents/BU/AN_588/AN588-Modules-nmer/country-data-2016.csv"
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactor = FALSE)
head(d)
```
Now use summary()

```{r Challenge1-2, echo=TRUE}
summary(d)
```
Let's check out the names of the columns

```{r Challenge1-3, echo=TRUE}
names(d)
```
#### What are the median **area** and **population size** of all countries in the data set? 

This is given in the summary so I can look back at the output and read it. 
The area median is 69700
The population median is 4.912e+06 (or 4,912,000)

To try pulling this more cleanly I can try the following using summary()

```{r Challenge1-4, echo=TRUE}
popsumm <- summary(d$population)
popsumm
areasumm <- summary(d[3])
areasumm
popsumm[3]
areasumm[3]
```

I used two different indexing/subsetting notations to create new objects that were then indexed again for just the median. 

Because the columns are named, the $ notation can be used before the name of the column which returns a 'summaryDefault' structure. When that is indexed for the third position (median), it returns a table. 

I used bracket notation for area summary which output a table. I indexed the table for the third position which returned a character vector.

I am not sure there is anything meaningful to these distinctions but might be useful in the future.

Now I will try using the median() function. na.rm is set to TRUE to remove the NA values from being considered.

```{r Challenge1-5, echo=TRUE}
median(d$population, na.rm = TRUE)
median(d$area, na.rm = TRUE)
```

Ok, this is way less clunky than the multiple functions I had to use for summary to make it as clean as this. The $ notation was the only one that worked for this one. [] notation did not work and returned an error "Error in median.default(d[3], na.rm = TRUE) : need numeric data"

#### Create a new pop_density variable in your data frame which is population / area. What are the 10 most dense countries? The 10 least dense?

First, I am using cbind() to add the pop_density variable as column and saving that into the d variable. I am using str() to ensure that the class of pop_density is numbers.

```{r Challenge1-6, echo=TRUE}
pop_density <- d$population/d$area
d <- cbind(d, pop_density)
head(d)
str(d)
```

Next, I am using order() to ifnd the top 10 most dense countries and saving them to a new variable d_topdense.

```{r Challenge1-7, echo=TRUE}
d_mostdense <- order(d[, "pop_density"], na.last = TRUE, decreasing = TRUE)
d_mostdense[1:10]
```

Doing this method only gives me the INDEXED POSITIONS of the countries (the row numbers) as a number vector. d_topdense is saved as an integer vector (just to prove a point). What I really want to see is a variable saved as a table or data frame so that I can look at the whole thing.

If I used order() inside index brackets I will be indexing just the ordered table.

```{r Challenge1-8, echo=TRUE}
d_mostdense <- d[order(d$pop_density, na.last = TRUE, decreasing = TRUE), ]
d_mostdense[1:10, ]
```

I had some errors (undefined columns selected) with this until I googled the error and found I needed to add a comma after the closing parenthesis and a space following that. I think this is because we are subsetting the rows in the first half of the brackets and we are ordering those by population density. We are not doing anything with the columns so that is left blank. R needs to know I am not doing anything with the columns even if that is leaving it blank with a comma and a space.

The same thing happens when I subset the variable d_mostdense for the first ten. Those are just the first ten rows, and I need to leave the column coordinate blank. If I do put in a coordinate for the column, it returns a integer vector of the values from the column (such as all of the density values but nothing else about them). Going back to the previous attempt at this, if I did add a column coordinate to that final line of code, I get an error saying incorrect number of dimensions. This is because d_mostdense was only an integer vector in that case and not a data frame, so it only has one dimension.

Attempting the same thing but with the least dense countries:

```{r Challenge1-9, echo=TRUE}
d_leastdense <- d[order(d$pop_density, na.last = TRUE, decreasing = FALSE), ]
d_leastdense[1:10, ]
```

#### Extract data from the 20 largest countries into a new variable. What are the median area and population size of these countries?

A country can be large by population or by area. I will do it by both I guess.

By area first:

```{r Challenge1-10, echo=TRUE}
d_bigarea <- d[order(d$area, na.last = TRUE, decreasing = TRUE), ]
d_bigarea <- d_bigarea[1:20, ]
d_bigarea
median(d_bigarea$population, na.rm = TRUE)
median(d_bigarea$area, na.rm = TRUE)
```

By population:

```{r Challenge1-11, echo=TRUE}
d_bigpop <- d[order(d$population, na.last = TRUE, decreasing = TRUE), ]
d_bigpop <- d_bigpop[1:20, ]
d_bigpop
median(d_bigpop$population, na.rm = TRUE)
median(d_bigpop$area, na.rm = TRUE)
```

#### Extract data from all countries beginning with the letters “A” through “F”. What are the mean area and population size of these countries?

Will need to use grep() for this which finds a pattern (in this case letters A through F) in a given character vector (in this case the column Countries). Note that note every column is a character class so this would not work on every column.

The extraction is similar to the order() extraction where the function is placed inside the bracket notation as the row dimension coordinate, and the column dimension is left blank.

```{r Challenge1-12, echo=TRUE}
alphabet <- d[grep("^[A-F]", d$country), ]
summary(alphabet)
mean(alphabet$population, na.rm = TRUE)
mean(alphabet$area, na.rm = TRUE)
```

Both summary() and mean() will get us the data requested which is mean area and population size of these 78 subset countries.

### Challenge 2

Using Boxplots and Barplots

Boxplots: boxplot() functions returns a box and whiskers visual of the summary data from a column plus the outliers. Does not like nonnumeric data so must be used with specific columns using bracket of $ notation

Barplots: barplot() function returns a bar plot with bar height proportional to the value of the variable. dotchart() provides similar. Barplots are good for crude data

#### Make boxplots of the raw population and area data, then do the same after log() transforming these variables.

```{r Challenge2-1, echo=TRUE}
boxplot(d$population)
boxplot(log(d$population))
boxplot(d$area)
boxplot(log(d$area))
```

Using log makes it so much more readable

Using par() with mfrow (filled in by row) as the argument will arrange the boxplots accordingly:

```{r Challenge2-2, echo=TRUE}
par(mfrow = c(2,2))
boxplot(d$population)
boxplot(log(d$population))
boxplot(d$area)
boxplot(log(d$area))
```

And now the barplots of population and area:

```{r Challenge2-2, echo=TRUE}
barplot(d$population)
barplot(d$area)
```

These plots do not mean much without labels.

### Challenge 3

#### Histograms 
The hist() function returns a histogram showing the complete empirical distribution of the data in binned categories, which is useful for checking skewness of the data, symmetry, multi-modality, etc. Setting the argument freq=FALSE will scale the Y axis to represent the proportion of observations falling into each bin rather than the count.

Make histograms of the log() transformed population and area data from the Country-Data-2016 file. Explore what happens if you set freq=FALSE versus the default of freq=TRUE. Try looking at other variables as well.

First attempt will just be with inputting the data to be turned into a histogram with no modifications to the variables.

```{r Challenge3-1, echo=TRUE}
par(mfrow = c(1, 2))
hist(log(d$population))
hist(log(d$area))
```

Frequency defaults to true, which is the y-axis. This is showing how frequent each population is and how frequent each area is (grouped up into bars that are equi-spaced by breaks). Breaks are another argument that can be modified.

If I use attach() then I don't need to use $ notation anymore and R will search just for the name of the column I am making a plot of

This time I will set freq to false which will make the y axis looking at density instead of frequency.I will also name the axes and the plot.

```{r Challenge3-2, echo=TRUE}
attach(d)
par(mfrow = c(1, 2))
hist(log(population), freq = FALSE, col = "orchid3", main = "Plot 1", xlab = "log(population size)", ylab = "density")
hist(log(area), freq = FALSE, col = "darkolivegreen4", main = "Plot 2", xlab = "log(area)", ylab = "density")
```

The y limits are not the same like they were with the freq=true histograms. That can be changed with an argument modification.

adding vertical lines showing mean of x axis.

```{r Challenge3-2, echo=TRUE}
par(mfrow = c(1, 2))
hist(log(population), freq = FALSE, col = "orchid3", main = "Plot 1", xlab = "log(population size)", ylab = "density", ylim = c(0, 0.2))
abline(v = mean(log(population), na.rm = TRUE), col = "hotpink")
hist(log(area), freq = FALSE, col = "darkolivegreen4", main = "Plot 2", xlab = "log(area)", ylab = "density", ylim = c(0, 0.2))
abline(v = mean(log(area), na.rm = TRUE), col = "hotpink")
```

#### Density Plot
The density() function computes a non-parametric estimate of the distribution of a variable, which can be combined with plot() to also yield a graphical view of the distribution of the data. If your data have missing values, then you need to add the argument na.rm=TRUE to the density() function. To superimpose a density() curve on a histogram, you can use the lines(density()) function.

```{r Challenge3-3, echo=TRUE}
par(mfrow = c(1, 1))  # set up one panel and redraw the log(population) histogram
hist(log(population), freq = FALSE, col = "lemonchiffon", main = "My Plot with Mean and Density",
    xlab = "log(population size)", ylab = "density", ylim = c(0, 0.2))
abline(v = mean(log(population), na.rm = TRUE), col = "hotpink")
lines(density(log(population), na.rm = TRUE), col = "aquamarine4")
detach(d)
```

### Challenge 4

Tables : the table() function can be used to summarize counts and proportions for categorical variables in your dataset.

Using the table() function, find what is the most common form of government in the Country-Data-2016 dataset. How many countries have that form? 

```{r Challenge4-1, echo=TRUE}
table(d$govt_form)
```

Using just the table function returns a list of all of the observations in the column requested and how many of each (e.g., there are 33 observations of constitutional monarchy)

```{r Challenge4-2, echo=TRUE}
sort(table(d$govt_form), decreasing = TRUE)
```

Remember that order() returns the indexed position (as seen earlier when I tried using it not inside bracket notation). Sort() returns the vector in order. 

**The most common form of government from the dataset is republic with 127 countries claiming that form.**

## Multivariate Data

### Challenge 5

Multiple boxplots or histograms can be laid out side-by-side or overlaid. For boxplots, the ~ operator can be read as “by”.

Read in the dataset KamilarAndCooperData, which contains a host of information from about 213 living primate species.

Spend some time exploring the data and then make boxplots of log(female body mass) ~ family. Try doing this with {base} graphics and then look at how we might do in in {ggplot2}, which provides a standard “grammar of graphics” (see the {ggplot2} documentation)

```{r Challenge5-1, echo=TRUE}
f1 <- "~/Documents/BU/AN_588/AN588-Modules-nmer/KamilarAndCooperData.csv"
k <- read.csv(f1, header = TRUE, sep = ",", stringsAsFactor = FALSE)
summary(k)
attach(k)
```
#### Boxplots
The column specified after the ~ goes on the X axis. This is non-numerical data otherwise it would not be able to be passed through the boxplot.
```{r Challenge5-2, echo=TRUE}
boxplot(log(Body_mass_female_mean) ~ Family)
detach(k)
```

The module instructions are: boxplot(log(Body_mass_female_mean) ~ Family, k) 
I do not see a difference adding the file variable at the end and I cannot find anything in the documentation for boxplot() for it.

Also in this plot not every family name is listed on the x axis.

#### ggplot2

With ggplot, we can add to the object little by little for ease of reading.

Here is a breakdown of the lines of code below
1. defining the variables
2. graph into a box plot
3. put x-axis names at 90deg to the axis
4. rename y axis
5. run the graph

```{r Challenge5-3, echo=TRUE}
p <- ggplot(data = k, aes(x = Family, y = log(Body_mass_female_mean)))
p <- p + geom_boxplot()
p <- p + theme(axis.text.x = element_text(angle = 90))
p <- p + ylab("log(Female Body Mass)")
p
```

### Challenge 6

Scatterplots with base vs ggplot2

Scatterplots : Scatterplots are a natural tool for visualizing two continuous variables and can be made easily with the plot(x=XXX, y=YYY) function in {base} graphics (where XXX* and YYY** denote the names of the two variables you wish to plot). Transformations of the variables, e.g., log or square-root (sqrt()), may be necessary for effective visualization.

Again using data from the KamilarAndCooperData dataset, plot the relationship between female body size and female brain size. Then, play with log transforming the data and plot again.

#### Base Graphics

```{r Challenge6-1, echo=TRUE}
attach(k)
par(mfrow = c(1, 2))
plot(x = Body_mass_female_mean, y = Brain_Size_Female_Mean)
plot(x = log(Body_mass_female_mean), y = log(Brain_Size_Female_Mean))
detach(k)
```

Did this to show the difference between no log and with log. With log is like zooming in and getting a better view of the data.

#### ggplot2

Following similar steps from the boxplot- will do a scatterplot with the same data as above
1. defining the variables and data and color coding the plot points by Family column
2. modify the axis labels (this was step 4 before)
3. make the scatterplot (step 2 before)
4. modify the legend at the bottom (kind of equivalent to step 3 before)
4. run the graph

```{r Challenge6-2, echo=TRUE}
p1 <- ggplot(data = k, aes(x = log(Body_mass_female_mean), y = log(Brain_Size_Female_Mean), color = factor(Family)))
p1 <- p1 + xlab("log(Female Body Mass)") + ylab("log(Female Brain Size)")
p1 <- p1 + geom_point()
p1 <- p1 + theme(legend.position = "bottom", legend.title = element_blank())
p1
```

Grid for faceting by a grouping variable

```{r Challenge6-3, echo=TRUE}
p1 <- p1 + facet_wrap(~Family, ncol = 4)
p1 <- p1 + theme(legend.position = "none")
p1
```

And then add regression lines

```{r Challenge6-4, echo=TRUE}
p1 <- p1 + geom_smooth(method = "lm", fullrange = TRUE)
p1
```

### Challenge 7