---
title: "Module-5"
author: "Nicole Merullo"
date: "2023-09-11"
output: html_document
---
# Loading Different Types of Plain Text Files
read.table() is the most basic way of bringing in a data file

## Reading from a Local File

```{r read file, echo=TRUE}
f <- file.choose() #CPDS-1960-2014-reduced.txt
d <- read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE, fill = T)
head(d) #lists the first 6 lines of data. Adding [1:10] would expand it to 10 etc.
```
Can also show the last 6 lines with the tail() command
```{r visualizing date, echo=TRUE}
tail(d)
```
```{r visualizing data, echo=TRUE}
class(d)
```

Alternatively we can visualize the file like this...

```{r read.delim, echo=TRUE}
d <- read.delim(f, header = TRUE, stringsAsFactors = FALSE)
head(d)
```
## Loading Comma-separated (.csv) text with {base} R functions
```{r csv files, echo=TRUE}
f <- file.choose()
```

```{r csv files 2, echo=TRUE}
d <- read.table(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

Or Alternatively

```{r csv files 3, echo=TRUE}
d <- read.csv(f, header = TRUE, stringsAsFactors = FALSE)
head(d)
```

#Using readr Package

```{r loading in readr, echo=TRUE}
require(readr)
```

```{r tsv file, echo=TRUE}
f <- file.choose() #CPDS-1960-2014-reduced.txt
d <- read_tsv(f, col_names = TRUE) #tsv is tab separated files
```
```{r tsv file 2, echo=TRUE}
head(d)
```

```{r tsv file 3, echo=TRUE}
class(d)
```

The class of this data set is both a data frame and other table based data structures

Alternatively we can load it in like this:

```{r tsv file 4, echo=TRUE}
d <- read_delim(f, delim = "\t", col_names = TRUE)
head(d)
```

Then readr with CSV

```{r csv file with readr, echo=TRUE}
require(readr)
f <- file.choose()
d <- read_csv(f, col_names = TRUE) 
```
```{r csv file with readr 2, echo=TRUE}
head(d) 
```

And alternatively:

```{r csv file with readr 3, echo=TRUE}
d <- read_delim(f, delim = ",", col_names = TRUE)
head(d)
```

## Loading Excel Files

First will try with readxl package
```{r readxl, echo=TRUE}
require(readxl)
f <- file.choose() #"~/Desktop/CPDS-1960-2014-reduced.xlsx"
d <- read_excel(f, sheet = 1, col_names = TRUE)
head(d)
str(d) #always use str() to check if your variables come in as the correct data class.
```

Now will try with XLConnect

```{r XLConnect, echo=TRUE}
require(XLConnect)
f <- file.choose() #"~/Desktop/CPDS-1960-2014-reduced.xlsx"
d <- readWorksheetFromFile(f, sheet = 1, header = TRUE)
head(d)
str(d)
```
 **This one cannot be completed because of missing java runtime.**
 If I am able to get this one to work - here is the rest of this part of XLConnect, which writes data frames back out to Excel worksheets:
 f <- "~/Desktop/output.xlsx"
if (file.exists(f)) {
    file.remove(f)
}
writeWorksheetToFile(f, d, sheet = "myData", clearSheets = TRUE)

## Loading Files Stored on a Remote Server
e.g., Github
Will use readr or curl for this

```{r curl, echo=TRUE}
library(curl)
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/CPDS-1960-2014-reduced.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

And for a tsv file, do this:

```{r curl for tsv, echo=TRUE}
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/CPDS-1960-2014-reduced.txt")
d <- read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
head(d)
```

And for readr csv do this:

```{r readr for csv, echo=TRUE}
Sys.setenv(VROOM_CONNECTION_SIZE = 131072 * 2)
library(readr)
f <- "https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/CPDS-1960-2014-reduced.csv"
d <- read_csv(f, col_names = TRUE)
head(d)
```

for readr tsv:

```{r readr for tsv, echo=TRUE}
f <- "https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/master/AN588_Fall23/CPDS-1960-2014-reduced.txt"
d <- read_tsv(f, col_names = TRUE)
head(d)
```

## Importing Data from a file hosted on Dropbox

```{r Dropbox commands, echo=FALSE}
require(rdrop2)
drop_auth()  # opens a browser dialog box to ask for authorization...
drop_dir()  # lists the contents of your dropbox folder
f <- "CPDS-1960-2014-reduced.csv"  # name of the file to read from
f <- drop_search(f)  # searches your dropbox directory for file or directory names; this can be slow
f <- f$path  # f$path is the location you choose from the results returned above
d <- drop_read_csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
str(d)
```

To read a file from someone else's dropbox do the following:

```{r Dropbox commands 2, echo=TRUE}
link <- "https://www.dropbox.com/scl/fi/560bh55pwv3sowbmyygsl/CPDS-1960-2014-reduced.csv?rlkey=sp4xkouegbno3fqrt65o7gb4d&dl=0"
link <- gsub(pattern = "dl=0", replacement = "dl=1", x = link)
d <- read.csv(link, header = TRUE, sep = ",", stringsAsFactors = FALSE) #you need to do this sub because the last characters of the link only bring you to the webpage, not to the raw data itself.
head(d)
str(d)
```

Another way to get data from a dropbox link:

```{r Dropbox with repmis, echo=TRUE}
require(repmis)
d <- source_data("https://www.dropbox.com/scl/fi/560bh55pwv3sowbmyygsl/CPDS-1960-2014-reduced.csv?rlkey=sp4xkouegbno3fqrt65o7gb4d&dl=1", header = TRUE, sep = ",")
head(d)
str(d)
```

##Loading in Files from Google Drive

This interactive interface will guide you through steps to link your Google Drive. The 'n_max' option will restrict the number of files retrieved to the 30 most recently-added to the linked drive. The 'type' option allows you to only search for a specific kind of file; in this case, spreadsheets (this 'type' will *not* find Excel files, only Google Sheets). It will store the results in an object called a 'dribble'.

In your dribble, find the Google Sheet you want. The first column is the name, and the second is the ID for the sheet. That ID is what you need to enter to read your google sheet into R (you may need to log in again to access it).

```{r googledrive, echo=TRUE}
require(googledrive)
require(googlesheets4)
f <- drive_find(n_max = 30, type = "spreadsheet") #reminder that you stored this as a variable called f!!
```

```{r googledrive 2, echo=TRUE}
d <- read_sheet("1Eo6UFxXnXKLhcZ9TXkoHl8GqIve_0ePRKNVb7xgP0L0")
head(d)
str(d)
```

##Downloading files from a remote server to your computer

rdrop2 and googledrive can take files from those remote servers and not just load them in, but also download them to your computer

Here is how to do it with dropbox, from your own dropbox account:

```{r dropbox download, echo=FALSE}
filename <- "CPDS-1960-2014-reduced.csv"  # name of file to download
f <- drop_search(filename)
f <- f$path  # f$path is the location of the results returned above
drop_get(f, local_file = paste0("~/Desktop/", filename), overwrite = TRUE, progress = TRUE)
```

This also works for other file types like excel and saves ti to the Desktop.

For doing this from Google Drive:

```{r google drive download, echo=FALSE}
drive_find(pattern = "CPDS-1960-2014-reduced")  #identifies files that match that name 
drive_download("CPDS-1960-2014-reduced.csv")
```

Instead, this method saves to the working directory.